\textsf{jaxoplanet}: Hardware-Accelerated Computation of Stellar Light Curves

### 2024-10-11 latex/paper.tex
big update

\title{Hardware-Accelerated Computation of Stellar Light Curves}

@@

\begin{abstract}
    The measurement of stellar fluxes over time serves a wide variety of science cases. Transit and occultation light curves, for example, are used to characterize the orbital parameters of stellar systems, and infer the atmospheric properties of their transiting bodies. Phase curves, on the other hand, can be used to study the thermal emission of exoplanets, and the atmospheric circulation of stars. In order to infer these properties, a fast, accurate and robust model of stellar light curves is needed, one that accounts for the non-uniform surface intensity of spherical bodies and their mutual occultations. The analytical framework known as \textsf{starry} satisfies these requirements, and was successfully applied over the years following its development. However, as datasets and model complexity grow, the tractable inference of these parameters becomes increasingly challenging, and often motivates biased approximations. In this paper, we present a new implementation of the \textsf{starry} framework in \texttt{jax}, a high-performance machine-learning library designed for large-scale computations. We describe the changes made to the original version of \textsf{starry} and evaluate the performance of its new implementation that we release in the open-source Python package \textsf{jaxoplanet}. Through this implementation, we provide the community with a differentiable model of stellar light curves, and enable the use of powerful machine-learning tools available in the \texttt{JAX} ecosystem. \gitlink{}

@@

During millenia, planets of the solar system were mostly known and studied through their orbits. Starting in the 16th century, Earth-based telescopes progressively enabled the direct imaging of their surface. But despite these new instruments, some minor planets were too small and too distant to be imaged, and remained unresolved until the 20th century, when more capable space-based observatories were launched. This is the case of Pluto, for example, whose surface was first studied through the light it reflects during its revolution and orbit around the Sun. Using point source light curves obtained from 1945 to 1986, \cite{Drish1995} tentatively reconstructed the map of the planet, only later compared to the first \textit{Hubble Space Telescope} low-resolution images obtained by \cite{Stern1997}. This mapping effort was gloriously completed with the extremely detailed pictures obtained during the close fly-by of the \textit{New Horizons} spacrecraft \citep{}.\\\\
The study of extrasolar planets is taking a path similar to that of Pluto, starting with the study of their unresolved light curves \citep{} while direct imaging instruments are being developed \citep{}. The same applies to the study of unresolved stellar atmospheres, whose properties are inferred from the light they emit. In this context, developments of accurate and robust models of extrasolar systems light curves are crucial. As these models are being compared to large datasets in probabilistic ways, their posterior likelihood must be evaluated millions of times, which require them to be fast. This inference can be highly facilitated by the advent of gradient-based optimization and inference tools, so that the differentiability of these models offer a clear advantage.\\\\
In this context, \cite{starry} presented an algorithm to analytically compute the light curves of spherical bodies with non-uniform surface intensities, particularly optimized for occultation light curves. This algorithm, known as \textit{starry}, has been successfully applied to a wide variety of science cases, from the study of exoplanets atmospheres \citep{} to the study of stellar photospheres \citep{}. However, as datasets and model complexity grow, the tractable inference of these parameters becomes increasingly challenging, and often motivates biased approximations. This is case, for example, in transmission spectroscopy measurements conducted with the \textit{James Webb Space Telescope}, for which ... \todo{(So, give an example of large number of spectral bins and required approximation) (Hatori et al. in prep)}.\\\\
In this paper, we present a new implementation of the \textit{starry} framework in \texttt{JAX}, a high-performance machine-learning library designed for large-scale computations. We start in \autoref{starry} with a summarry of the formalism introduced by \cite{starry}. In \autoref{optimization}, we describe the changes made to the original version of \textit{starry}. Finally, in \autoref{performance}, we evaluate the performance of our new implementation that we release in the open-source Python package \textsf{jaxoplanet}. Through this implementation, we provide the community with a differentiable model of stellar light curves, and enable the use of powerful machine-learning tools available in the \texttt{JAX} ecosystem.\\\\

@@

\newpage
\section{Light curve models}\label{starry}

Inspired by the idea of \cite{pal2012}, \cite{starry} and \cite{Agol2020} present a general framework to compute the light curves of spherical bodies with non-uniform surface intensities. This framework, known as \textit{starry}, allows for precise and efficient computation of the light curves of stars and exoplanets, and has been successfully applied to a wide variety of science cases. In this section, we describe the mathematical formalism behind the \textit{starry} model.

@@

The surface of a spherical body can be naturally described by spherical harmonics: a set of special functions defined on the sphere that can serve as orthonormal basis to represent any spherical surface map. If $\bvec{\tilde y}$ denotes the spherical harmonics basis of degree $l_{max}$ (defined in \citealt[section X]{starry}) and $\bvec{y}$ a vector that contains the coefficients describing the surface in this basis, the intensity $I$ of the map projected onto the sky-projected $(x, y)$ plane of the observer is

@@

In order to simplify this integral, \citealt{starry} introduce a change of basis from the spherical harmonics basis $\bvec{\tilde y}$ to a simple polynomial basis $\bvec{\tilde p}$ characterized by the change of basis matrix $\bvec{A_1}$. In this new basis, $\bvec{y}$ can be expressed as

@@

The same formalism can be used to compute the integrated intensity of a surface map occulted by another spherical body. In this case, the intensity of the map must be integrated only over the unocculted surface $S$ of the disk, such that 

@@

To simplify this expression, \citealt{pal2012} note that any two-dimensional integral of the unocculted disk can be reduced to a one-dimensional integral over the contour $C$ of the occulting body using Green's theorem\footnote{Also known as Stokes' theorem in three-dimensional vector calculus.}. This can be done if the integrand of the integral over S can be described by a function $f$ such that

@@

By applying a last change of basis matrix $\bvec{A_2}$ to the polynomial basis $\bvec{\tilde p}$, \citealt{starry} transforms $\bvec{\tilde p}$ to the \textit{Green's basis} $\bm{\tilde g}$ that satisfies this property, meaning that the surface integral of each of its components $\tilde{g_n}$ can be transformed to the line integral

@@

\subsection{Limb-darkened surfaces}
We notice that a simpler situation is encountered when the surface of the spherical body can be described by a radial intensity profile $I(r)$, such as a limb-darkened star. In this case, the vector of spherical harmonics coefficients $\bvec{y}$ is very sparse (because all $m\neq0$ components are zero) and the surface map doesn't need to be rotated. Contemporary to \cite{starry}, \cite{Agol2020} explore this case and provide simple expressions for the integrated intensity of a star whose surface can be described by a polynomial limb darkening law occulted by an opaque body. In this case:
\begin{enumerate}
    \item The surface is simply described by $\bvec{u}$, a vector of polynomial limb darkening coefficients.
    \item The Green's basis takes a simpler form (see \citealt[Eq. X]{Agol2020}).
    \item Matrices involved in the computation of the integrated flux have $l_{max}$ columns, versus $(l_{max}+1)^2$ in the general case, reducing drastically the computational cost of the model.
\end{enumerate}
For these reasons, and because of its wide application, our implementation treats radially-symmetric maps separately, leading to highly optimize computations of limb-darkened light curves.\\\\

\newpage
\section{Implementation details}\label{optimization}

\subsection{Limb darkened maps}
Limb darkened maps are obtained by multiplying a base map, devoid of limb darkening, with a pure limb darkened map. If $\bvec{U}$ is the change of basis matrix that transforms the polynomial limb darkening coefficients $\bvec{u}$ to the spherical harmonics coefficients $\bvec{y}_u$, the pure limb darkened map is
\begin{equation}
    \label{eq:limb_darkening_map}
    \bvec{y}_u = \bvec{U}\bvec{u}.
\end{equation}
The multiplication of the base map represented by $\bvec{y}$ with the limb darkening map represented by $\bvec{y}_u$ can be seen as the multiplication of two polynomials, only with special coefficients (called Clebsch–Gordan coefficients) inherent to the properties of spherical harmonics. To simplify this operation, we transform both maps into the polynomial basis, so that their product is the actual product of two polynomials which doesn't need the computation of special coefficients.

@@

As shown in \autoref{starry}, computing the flux of the surface at a given time $t$ involves a rotation of the spherical harmonics basis, from the rest-frame of the star (\autoref{fig:rotation_basis}, left) to its sky-projected orientation (\autoref{fig:rotation_basis}, right), with a final rotation relative to the position of the occulting body (see Figure 2. from \citealt{starry}).
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.6\textwidth]{../workflows/rotations/figures/rotation_basis.pdf}
        \caption{Rotation of a primary body's surface to place it in the sky-projected frame of the observer.}
        \label{fig:rotation_basis}
    \end{center}
\end{figure}

The rotation of spherical harmonics involves the computation of Wigner-D-matrices, commonly obtained through robust recursion relations in both degree $l$ and order $m$, but leading to costly computations. Hence, knowing how to decompose the full spherical harmonics rotation using pre-computed rotations is essential to achieve optimal performance in this framework. \autoref{fig:rotation_starry} shows the 6 elementary rotations currently used in \textsf{starry}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/rotations/figures/rotation_starry.pdf}
        \caption{Decomposition of the full rotation of the surface into elementary rotations. Each schematic shows a sphere rotated from a previous orientation around an axis displayed as a black vector and an angle shown on top.}
        \label{fig:rotation_starry}
    \end{center}
\end{figure}
In this figure, the first 3 rotations have the effect to rotate the star around its rotation axis, and rotations 3 to 5 place the star on a sky-projected frame. Finally, rotation 6 aligns the spherical harmonics map to the occulting body on the $y$ axis in order to apply Green's theorem in the appropriate basis (Figure 2 from \citealt{starry}).\\\\
The first thing to notice is that rotations 1 to 3 could be simply reduced to a rotation of angle $\theta$ about the y-axis, turning steps 1 to 3 into one. However, rotations around the pole, i.e.\,with a rotation axis along $z$, have simpler expressions that can be implemented separately. In addition, pre-computing these matrices at lower cost is particularly important for steps 2 and 6, involving a potentially large set of angles $\{\theta\}$ and $\{\theta_z\}$, defined over times $\{t\}$, for which to compute the full light curve. This explains the current decomposition of the complete rotation in six separate steps.\\\\
In \textsf{jaxoplanet}, we compute the Wigner-D-matrices by employing the Risbo recursion relations \citep{Risbo1996} implemented in \texttt{JAX} as part of the \texttt{s2fft} Python package \citep{price:s2fft}. In addition, we merge rotations 3, 4 and 5 (see \autoref{fig:rotation_jaxoplanet}) into a single compound rotation of axis
\begin{equation}
    \bvec{v} = \frac{1}{\sqrt{1 - \cos^2{\left(\frac{inc}{2} \right)} \cos^2{\left(\frac{obl}{2}\right)}}} \begin{pmatrix}
        \sin{\left(\frac{inc}{2} \right)} \cos{\left(\frac{obl}{2}\right)}\\
        \sin{\left(\frac{inc}{2} \right)} \sin{\left(\frac{obl}{2}\right)}\\
         - \cos{\left(\frac{inc}{2} \right)} \sin{\left(\frac{obl}{2}\right)}\\
    \end{pmatrix},
\end{equation}
and angle
\begin{equation}
    \label{eq:combined_angle}
    \omega = 2 \cos^{-1}{\left(\cos{\left(\frac{inc}{2} \right)} \cos{\left(\frac{obl}{2}\right)} \right)}.
\end{equation}
This way, the complete rotation reduces to the four separate steps shown in \autoref{fig:rotation_jaxoplanet}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/rotations/figures/rotation_jaxoplanet_1.pdf}
        \caption{Consecutive rotations of the spherical harmonics basis in the \textsf{jaxoplanet} implementation. Each schematic shows a sphere rotated from a previous orientation around an axis displayed as a black vector and an angle shown on top.}
        \label{fig:rotation_jaxoplanet}
    \end{center}
\end{figure}

@@

We described in \autoref{occultation} how the integrated intensity of a spherical body occulted by another can be computed using the solution vector $\bvec{s}^\mathsf{T}$. \cite{pal2012} emphasizes that, knowing $\bvec{G}_n$, each component $s_n$ of the solution vector defined in \autoref{eq:sn_gn} can be decomposed into
\begin{equation}
    s_n = \mathcal{Q}(\bvec{G}_n) - \mathcal{P}(\bvec{G}_n)
\end{equation}
where $\mathcal{Q}(\bvec{G}_n)$ and $\mathcal{P}(\bvec{G}_n)$ are line integrals over the contours of the occulting body and the primary's surface, as shown in \autoref{fig:geometry}.
%
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/geometry/figures/occultation_geometry.pdf}
        \caption{Geometric configuration and angles definitions for the two-body occultation. On the left figure, the larger light grey disk corresponds to the primary body being occulted by the darker secondary body. On the right figure, the area $S$ on which the intensity is integrated is shown in light blue. By choosing an appropriate transform, the integration over $S$ can be reduced to a line integral over the contour of the occulting body, which can be decomposed into the contour of the occulting body within the area of the primary body (in cyan) and the contour of the primary body everywhere else (in blue).}
        \label{fig:geometry}
    \end{center}
\end{figure}
%
In the general case, the $\mathcal{Q}$ integral takes a simple analytical form and can be computed using the recurrence relations from \cite{starry} (Equation D29). The case of a limb-darkened surface is even simpler as $\mathcal{Q}(\bvec{G}_n) = 0$ for all $n > 0$. On the other hand, the $\mathcal{P}$ integral is harder to evaluate and requires the use of recurrence relations and truncated series expansions, with a scheme that depends on the geometric configuration of the system (see \citealt{starry}, Section D.2).\\\\
In order to translate the \textsf{starry}  implementation to \textsf{JAX}, we evaluate the $\mathcal{P}$ integrals numerically using Gauss-Legendre integrations, requiring the evaluation of the integrated functions only over a fixed number of points.\\\\

\subsubsection{Polynomial limb darkening}
For a surface described by a polynomial limb darkening law of order $n$, the $\mathcal{P}$ integral is given in \cite{Agol2020} by
\begingroup\makeatletter\def\f@size{9}\check@mathfonts
$$
    \mbox{\normalsize$\mathcal{P}(\bvec{G}_n)$} =
    \begin{dcases}
        %
        \pi - \kappa_1 - r^2\kappa_0 + \textit{A}_{kite} & \qquad n = 0
        \\[1em]
        2s_0 + 4\pi\eta - 2\pi & \qquad n = 2
        \\[1em]
        2r(4br)^{\frac{n}{2}}
        \int\displaylimits_{-\kappa_0/2}^{\kappa_0/2}
        (k^2 - \sin^2{\varphi '})^{\frac{n}{2}}
        (r - b + 2b\sin^2\varphi ') \, \text{d}\varphi ' & \qquad \mathrm{n\ge3},
    \end{dcases}
$$
\endgroup
where $\kappa_0$, $\kappa_1$ and $A_{kite}$ are the angles and area shown in \autoref{fig:geometry} and computed following \cite{Agol2020} (Equations 31 and 32). For $n=2$, $\eta$ is defined in \cite{Agol2020} (Equation 53) and depends on the elliptic parameter
%
\begin{equation}
    k = \frac{1-r^2 - b^2 + 2br}{4br}.
\end{equation}\\
%
For $n=1$, we use Equation 31, 32 and 34 from \cite{starry} to write
%
\begin{equation}
    \mathcal{P}(\bvec{G}_1) = \int\displaylimits_{-\kappa_0/2}^{\kappa_0/2}
        r(r-b\cos{2\varphi '})\;\frac{1-z^3}{3(1-z^2)}\text{d}\varphi ' \quad \text{and} \quad \mathcal{Q}(\bvec{G}_1) = \frac{2}{3}(\pi-\kappa_1),
\end{equation}\\
%
where $1 - z^2 = b + r - 2\cos{\varphi '}$ \footnote{note that $\varphi '$ is different from the angle $\varphi$ shown in \autoref{fig:geometry} and was introduced through the change of variable $\varphi ' = \frac{1}{2}(\varphi -\frac{3\pi}{2})$.}.
\subsubsection{Non-uniform surface}
In the general case, for a star with a non-radially symmetric surface intensity, $n$ denotes the index of the spherical harmonics component such that
\begin{equation}
    n = l^2 + l + m,
\end{equation}
where $l$ is the degree of the spherical harmonic and $m$ its order. For each component, the expression of $\mathcal{P}(\bvec{G}_n)$ depends on $\mu$ and $\nu$ defined as
\begin{equation}
    \begin{aligned}
    \mu &\equiv l - m,\\
    \nu &\equiv l + m.
    \end{aligned}
\end{equation}

@@

>>>>

\title{\textsf{jaxoplanet}: Hardware-Accelerated Computation of Stellar Light Curves}

@@

\begin{abstract}
    Modeling light curves of extrasolar systems serves a wide variety of science cases. Transit and occultation light curves, for example, are used to characterize their orbital configuration, and infer the atmospheric properties of their transiting bodies. Phase curves, on the other hand, can be used to study the thermal emission of exoplanets, and the atmospheric circulation of stars. In order to infer these properties, a fast, accurate and robust model of stellar light curves is needed, one that accounts for the non-uniform surface intensity of spherical bodies and their mutual occultations. In this context, the analytical framework known as \textsf{starry} provided the community with a powerful tool, instrumental in .... However, as datasets and models complexity grow, the tractable inference of these parameters becomes increasingly challenging, and often motivates biased approximations. In this paper, we present a new implementation of the \textsf{starry} framework in \texttt{jax}, a high-performance machine-learning library designed for large-scale computations. We describe the changes made to the original version of \textsf{starry} and evaluate the performance of its new implementation that we release in the open-source Python package \textsf{jaxoplanet}. Through this implementation, we provide the community with a differentiable model of stellar light curves, and enable the use of powerful machine-learning tools available in the \textsf{JAX} ecosystem. \gitlink{}

@@

Planets of the solar system used to be known and studied through their position in the sky. It is only starting in the 16th century that Earth-based telescopes progressively enabled the direct imaging of their surface. But despite these new instruments, some minor planets were too small and too distant to be imaged, and remained unresolved until the end of the 20th century, when more capable space-based observatories were launched. This is the case of Pluto, for example, whose surface was first studied thanks to the light it reflects while rotating and orbiting the Sun. Using point source observations obtained from 1945 to 1986, \cite{Drish1995} tentatively reconstructed the map of the planet, and only later compared it to the first \textit{Hubble Space Telescope} low-resolution images obtained by \cite{Stern1997}. This mapping effort was gloriously completed with the extremely detailed pictures obtained by the \textit{New Horizons} spacrecraft during its fly-by of the planet \citep{Stern2015}.\\\\
Although imaging the surface of an exoplanet might be several decades away, their study follows that of Pluto, by first employing indirect methods and unresolved light curves (see \citealt{Mayor1995} and \citealt{Charbonneau2000})  while waiting for the development of more powerful direct imaging instruments \cite{}. Similarly, stellar photospheres can currently be mapped using interferometry and Doppler imaging, but these techniques are costly and can only be applied to a handful of objects, so that time-resolved spectrophotometry remains our main way to study them. In this context, the development of accurate and robust models of stellar light curves is crucial.\\\\
In the age of wide-field surveys (such as) and high-precision spectroscopic instruments, these models are being compared to large datasets in probabilistic ways, requiring fast and robust evaluations of their posterior likelihood over larger and larger parameter spaces. These inferences can be highly facilitated by gradient-based optimization and sampling algorithms, so that being able to differentiate these models with respect to their parameters offers clear advantages.\\\\
In this context, \cite{starry} and \cite{Agol2020} presented a framework to analytically compute the light curves of spherical bodies with non-uniform surface intensities, such as stars and exoplanets, and the flux observed from their mutual occultations. This framework, known as \textit{starry}, has been successfully applied to a wide variety of science cases, like the study of transiting exoplanets \citep{Kostov2019}, precise modeling of radial velocities \citep{Dawson2021}, the study of exoplanets atmospheres \citep{Bell2023} or the study of stellar photospheres \citep{Almenara2022}. However, as datasets and models complexity grow, the tractable inference of the larger parameter spaces we have to deal with becomes increasingly challenging, and often motivates biased approximations. This is the case, for example, in transmission spectroscopy measurements conducted with the \textit{James Webb Space Telescope} NIRSpec instrument, that should ideally consist in the simultaneous modeling of more than 3000 transit light curves, one per pixel element (e.g. \citealt{Alderson2023}).\\\\
In this paper we present \textsf{jaxoplanet}, a new implementation of the \textit{starry} framework in \texttt{JAX}, a high-performance machine-learning library designed for large-scale computations, including autodifferentiation capabilities, deployed on various hardware. In \autoref{starry}, we summarize the formalism introduced by \cite{starry} to analytically compute stellar light curves. In \autoref{optimization}, we describe the \textsf{jaxoplanet} implementation and the changes made to the \textsf{starry} implementation. Finally, in \autoref{performance}, we evaluate the performance of our new implementation that we release in the open-source Python package \textsf{jaxoplanet}. Through this implementation, we provide the community with a differentiable model of stellar light curves, and enable the use of powerful machine-learning tools available in the \texttt{JAX} ecosystem.\\\\

@@

\section{Light curve models}\label{starry}

Inspired by the idea of \cite{pal2012}, \cite{starry} and \cite{Agol2020} present a general framework to compute the light curves of spherical bodies with non-uniform surface intensities. In this section, we describe the mathematical formalism behind the \textit{starry} model.

@@

The surface of a spherical body can be naturally described by spherical harmonics: a set of special functions defined on the sphere that can serve as orthonormal basis to represent any spherical surface map. If $\bvec{\tilde y}$ denotes the spherical harmonics basis of degree $l_{max}$ (defined in \citealt[section 2.2]{starry}) and $\bvec{y}$ a vector that contains the coefficients describing the surface in this basis, the intensity $I$ of the map projected onto the sky-projected $(x, y)$ plane of the observer is

@@

In order to simplify this integral, \citealt{starry} introduce a change of basis from the spherical harmonics basis $\bvec{\tilde y}$ to a simple polynomial basis $\bvec{\tilde p}$ (defined in \citealt[section 2.3]{starry}) characterized by the change of basis matrix $\bvec{A_1}$. In this new basis, $\bvec{y}$ can be expressed as

@@

In order to model transits and occultations, the same formalism can be used to compute the integrated intensity of a surface map occulted by another spherical body. In this case, the intensity of the map must be integrated only over the unocculted surface $S$ of the disk, such that 

@@

To simplify this expression, \cite{pal2012} note that any two-dimensional integral of the unocculted disk can be reduced to a one-dimensional integral over the contour $C$ of the occulting body using Green's theorem\footnote{Also known as Stokes' theorem in three-dimensional vector calculus.}. This can be done if the integrand of the integral over S can be described by a function $f$ such that

@@

By applying a last change of basis matrix $\bvec{A_2}$ to the polynomial basis $\bvec{\tilde p}$, \cite{starry} transforms $\bvec{\tilde p}$ to the \textit{Green's basis} $\bm{\tilde g}$ that satisfies this property, meaning that the surface integral of each of its components $\tilde{g_n}$ can be transformed to the line integral

@@

\subsection{Limb-darkened surfaces}
We notice that a simpler situation is encountered when the surface of the spherical body can be described by a radial intensity profile $I(r)$, such as a limb-darkened star. In this case, the vector of spherical harmonics coefficients $\bvec{y}$ is very sparse (because all $m\neq0$ components are zero) and the surface map doesn't need to be rotated. Contemporary to \cite{starry}, \cite{Agol2020} explore this case and provide simple expressions for the integrated intensity of a star whose surface can be described by a polynomial limb darkening law occulted by an opaque body. In this case:
\begin{enumerate}
    \item The surface is simply described by $\bvec{u}$, a vector of polynomial limb darkening coefficients.
    \item The Green's basis takes a simpler form (see \citealt[Equation 14]{Agol2020}).
    \item Matrices involved in the computation of the integrated flux have $l_{max}$ columns, versus $(l_{max}+1)^2$ in the general case, reducing drastically the computational cost of the model.
\end{enumerate}
For these reasons, and because of its wide application, our implementation treats radially-symmetric maps separately, leading to highly optimize computations of limb-darkened light curves.\\\\

\newpage
\section{Implementation details}\label{optimization}
The formalism introduced by \cite{starry}, summarized in \autoref{starry}, provides an elegant analytical model of stellar light curves. Implemented in C++ and released in the \texttt{starry} Python package, it was built for the probabilistic programming library \textsf{PyMC3} and led to several advances in the field of exoplanetary science. However, since its initial development, the \textsf{PyMC} project adopted several breaking changes that made \textsf{starry} incompatible with its latest versions. Updating \textsf{starry} to the latest version of \textsf{PyMC} would require a complete rewriting of the code, that would again be highly dependent on the latter. Hence, our first motivation comes from the broken dependencies of \textsf{starry} and aims to provide the community with an implementation compatible with a wider ecosystem of tools. In this respect \textsf{JAX} represents an interesting alternative. This high-performance machine-learning library can be used to perform hardware-accelerated computations on various type of devices (CPU, GPU, TPU) and gave rise to probabilistic tools developped by an ever-growing community of scientists and engineers. However, \textsf{JAX} capabilities really shine when models are written with a functional-programming mindset, and computations performed in a vectorized way.\\\\
In this section, we describe the changes made to the original \textsf{starry} implementation to allow its rewriting in \textsf{JAX}. We also highlight details about the latest version of \textsf{starry}'s original implementation that were omitted in \citealt{starry} but described in parts of its code and documentation.

\subsection{Limb darkened maps}
Limb darkened maps are obtained by multiplying a base map, devoid of limb darkening, with a pure limb darkened map. If $\bvec{U}$ is the change of basis matrix that transforms the polynomial limb darkening coefficients $\bvec{u}$ to the spherical harmonics coefficients $\bvec{y}_u$, the pure limb darkened map is
\begin{equation}
    \label{eq:limb_darkening_map}
    \bvec{y}_u = \bvec{U}\bvec{u}.
\end{equation}
The multiplication of the base map represented by $\bvec{y}$ with the limb darkening map represented by $\bvec{y}_u$ can be seen as the multiplication of two polynomials, only with special coefficients (called Clebsch–Gordan coefficients) inherent to the properties of spherical harmonics. To simplify this operation, we transform both maps into the polynomial basis, so that their product is the actual product of two polynomials which doesn't require the computation of special coefficients.

@@

As shown in \autoref{starry}, computing the flux of the surface at a given time $t$ involves a rotation of the spherical harmonics basis, from the rest-frame of the star (\autoref{fig:rotation_basis}, left) to its sky-projected orientation (\autoref{fig:rotation_basis}, right), with a final rotation relative to the position of the occulting body (see Figure 2. from \citealt{starry}).
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.6\textwidth]{../workflows/geometry/figures/rotation_basis.pdf}
        \caption{Rotation of a primary body's surface to place it in the sky-projected frame of the observer. \codelink{workflows/geometry/scripts/rotations.ipynb}}
        \label{fig:rotation_basis}
    \end{center}
\end{figure}

The rotation of spherical harmonics involves the computation of Wigner-D-matrices, commonly obtained through robust recursion relations in both degree $l$ and order $m$, but leading to costly computations. Hence, knowing how to decompose the full spherical harmonics rotation using pre-computed rotations is essential to achieve optimal performance in this framework. \autoref{fig:rotation_starry} shows the 6 elementary rotations currently used in \textsf{starry}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/geometry/figures/rotation_starry.pdf}
        \caption{Decomposition of the full rotation of the surface into elementary rotations. Each schematic shows a sphere rotated from a previous orientation around an axis displayed as a black vector and an angle shown on top. \codelink{workflows/geometry/scripts/rotations.ipynb}}
        \label{fig:rotation_starry}
    \end{center}
\end{figure}
In this figure, the first 3 rotations have the effect to rotate the star around its rotation axis, and rotations 3 to 5 place the star on a sky-projected frame. Finally, rotation 6 aligns the spherical harmonics map to the occulting body on the $y$ axis in order to apply Green's theorem in the appropriate basis (Figure 2 from \citealt{starry}).\\\\
The first thing to notice is that rotations 1 to 3 could be simply reduced to a rotation of angle $\theta$ about the y-axis, turning steps 1 to 3 into one. However, rotations around the pole, i.e.\,with a rotation axis along $z$, have simpler expressions that can be implemented separately. In addition, pre-computing these matrices at lower cost is particularly important for steps 2 and 6, involving a potentially large set of angles $\{\theta\}$ and $\{\theta_z\}$, defined over times $\{t\}$, for which to compute the full light curve. This explains the current decomposition of the complete rotation in six separate steps.\\\\
In \textsf{jaxoplanet}, we compute the Wigner-D-matrices by employing the Risbo recursion relations \citep{Risbo1996} implemented in \texttt{JAX} as part of the \texttt{s2fft} Python package \citep{price:s2fft}. In addition, we merge rotations 3, 4 and 5 (see \autoref{fig:rotation_jaxoplanet}) into a single compound rotation of axis
\begin{equation}
    \bvec{v} = \frac{1}{\sqrt{1 - \cos^2{\left(\frac{inc}{2} \right)} \cos^2{\left(\frac{obl}{2}\right)}}} \begin{pmatrix}
        \sin{\left(\frac{inc}{2} \right)} \cos{\left(\frac{obl}{2}\right)}\\
        \sin{\left(\frac{inc}{2} \right)} \sin{\left(\frac{obl}{2}\right)}\\
         - \cos{\left(\frac{inc}{2} \right)} \sin{\left(\frac{obl}{2}\right)}\\
    \end{pmatrix},
\end{equation}
and angle
\begin{equation}
    \label{eq:combined_angle}
    \omega = 2 \cos^{-1}{\left(\cos{\left(\frac{inc}{2} \right)} \cos{\left(\frac{obl}{2}\right)} \right)}.
\end{equation}
This way, the complete rotation reduces to the four separate steps shown in \autoref{fig:rotation_jaxoplanet}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/geometry/figures/rotation_jaxoplanet_1.pdf}
        \caption{Consecutive rotations of the spherical harmonics basis in the \textsf{jaxoplanet} implementation. Each schematic shows a sphere rotated from a previous orientation around an axis displayed as a black vector and an angle shown on top. \codelink{workflows/geometry/scripts/rotations.ipynb}}
        \label{fig:rotation_jaxoplanet}
    \end{center}
\end{figure}

@@

We described in \autoref{occultation} how the integrated intensity of a spherical body occulted by another can be computed using the solution vector $\bvec{s}^\mathsf{T}$. \cite{pal2012} emphasizes that, knowing $\bvec{G}_n$, each component $s_n$ of the solution vector defined in \autoref{eq:sn_gn} can be decomposed into
\begin{equation}
    s_n = \mathcal{Q}(\bvec{G}_n) - \mathcal{P}(\bvec{G}_n)
\end{equation}
where $\mathcal{Q}(\bvec{G}_n)$ and $\mathcal{P}(\bvec{G}_n)$ are line integrals over the contours of the occulting body and the primary's surface, as shown in \autoref{fig:geometry}.
%
\begin{figure}[H]
    \begin{center}
        \includegraphics[width=\textwidth]{../workflows/geometry/figures/occultation_geometry.pdf}
        \caption{Geometric configuration and angles definitions for the two-body occultation. On the left figure, the larger light grey disk corresponds to the primary body being occulted by the darker secondary body. On the right figure, the area $S$ over which the intensity is integrated is shown in light blue. By choosing an appropriate transform, the integration over $S$ can be reduced to a line integral over the contour of the occulting body, which can be decomposed into the contour of the occulting body within the area of the primary body (in cyan) and the contour of the primary body everywhere else (in blue). \codelink{workflows/geometry/scripts/occultation_geometry.ipynb}}
        \label{fig:geometry}
    \end{center}
\end{figure}
%
In the general case, the $\mathcal{Q}$ integral (over the blue contour shown in \autoref{fig:geometry}) takes a simple analytical form and can be computed using the recurrence relations from \cite{starry} (Equation D29). The case of a limb-darkened surface is even simpler as $\mathcal{Q}(\bvec{G}_n) = 0$ for all $n > 0$. On the other hand, the $\mathcal{P}$ integral (over the cyan contour shown in \autoref{fig:geometry}) is harder to evaluate and requires the use of recurrence relations and truncated series expansions, with a scheme that depends on the geometric configuration of the system (see \citealt{starry}, Section D.2).\\\\
In order to translate the \textsf{starry}  implementation to \textsf{JAX}, we evaluate the $\mathcal{P}$ integrals numerically using Gauss-Legendre integrations, requiring the evaluation of the integrated functions only over a fixed number of points $q$.\\\\

\subsubsection{Polynomial limb darkening}
For a surface described by a polynomial limb darkening law of order $n$, the $\mathcal{P}$ integral is given in \cite{Agol2020} by
\begingroup\makeatletter\def\f@size{9}\check@mathfonts
$$
    \mbox{\normalsize$\mathcal{P}(\bvec{G}_n)$} =
    \begin{dcases}
        %
        \pi - \kappa_1 - r^2\kappa_0 + \textit{A}_{kite} & \qquad n = 0
        \\[1em]
        \text{see below} & \qquad n = 1
        \\[1em]
        2s_0 + 4\pi\eta - 2\pi & \qquad n = 2
        \\[1em]
        2r(4br)^{\frac{n}{2}}
        \int\displaylimits_{-\kappa_0/2}^{\kappa_0/2}
        (k^2 - \sin^2{\varphi '})^{\frac{n}{2}}
        (r - b + 2b\sin^2\varphi ') \, \text{d}\varphi ' & \qquad \mathrm{n\ge3},
    \end{dcases}
$$
\endgroup
where $\kappa_0$, $\kappa_1$ and $A_{kite}$ are the angles and area shown in \autoref{fig:geometry} and computed following \cite{Agol2020} (Equations 31 and 32). For $n=2$, $\eta$ is defined in \cite{Agol2020} (Equation 53) and depends on the elliptic parameter
%
\begin{equation}
    k = \frac{1-r^2 - b^2 + 2br}{4br},
\end{equation}\\
%
also used in the case $n=3$. For $n=1$, we use Equation 31, 32 and 34 from \cite{starry} to write
%
\begin{equation}
    \mathcal{P}(\bvec{G}_1) = \int\displaylimits_{-\kappa_0/2}^{\kappa_0/2}
        r(r-b\cos{2\varphi '})\;\frac{1-z^3}{3(1-z^2)}\text{d}\varphi ' \quad \text{and} \quad \mathcal{Q}(\bvec{G}_1) = \frac{2}{3}(\pi-\kappa_1),
\end{equation}\\
%
where $1 - z^2 = b + r - 2\cos{\varphi '}$ \footnote{note that $\varphi '$ is different from the angle $\varphi$ shown in \autoref{fig:geometry} and was introduced through the change of variable $\varphi ' = \frac{1}{2}(\varphi -\frac{3\pi}{2})$.}.
\subsubsection{Non-uniform surface}
In the general case, for a star with a non-radially symmetric surface intensity, $n$ denotes the index of the spherical harmonics component such that
\begin{equation}
    n = l^2 + l + m,
\end{equation}
where $l$ is the degree of the spherical harmonic and $m$ its order. For each component, the expression of $\mathcal{P}(\bvec{G}_n)$ depends on $\mu$ and $\nu$ defined as
\begin{equation}
    \begin{aligned}
    \mu &\equiv l - m,\\
    \nu &\equiv l + m.
    \end{aligned}
\end{equation}

@@

In addition, \autoref{fig:precision_degree} shows the evolution of the error on $f$ for increasing degrees of the spherical harmonics for \textsf{starry} and \textsf{jaxoplanet}. With these results, we validate the precision of \textsf{jaxoplanet} and show it is on per with the legacy C++ \textsf{starry} implentation\footnote{version 1.2.0 \citep{starry_120}} described in \cite{starry}.
